Forest canopies are better generators of biodiversity in the tropics

### 2025-07-17 manuscript.Rmd
revise ms methods - microclimate methods

---
bibliography: "`r rbbt::bbt_write_bib('r-refs.bib', overwrite = TRUE)`"
csl: science.csl
output: 
  officedown::rdocx_document:
    reference_docx: "style-guide.docx"
    mapstyles:
      Normal: ['First Paragraph']
plots:
  caption:
    style: Image Caption
    pre: 'Figure '
    sep: ': '
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \rhead{Global Verticality}
- \begin{document}
---

@@

Maximum and minimum temperatures had less consistent effects across taxa (Fig. \@ref(fig:fig3)), but
may still place physiological constraints on arboreal life strategies that influence the capacity
for taller canopies to support richer species assemblages, particularly in regions exhibiting
steeper vertical gradients characterized by hotter maximum and colder minimum temperatures. For
example, high maximum temperatures generally reduced verticality and/or reduced the positive impact
of canopy height on verticality, barring mammals for which verticality responded positively to
maximum temperatures. High temperatures may cause ectotherm body temperatures to exceed critical
thermal limits [@sundayThermalsafetyMarginsNecessity2014]. While consistent water availability in
the canopies of tropical wet forests may allow ectotherms to maintain body temperature below
critical limits, the drier conditions of temperate forests combined with steeper vertical gradients
in maximum temperatures and canopy temperatures that exceed 45°C may impede the use of arboreal
habitats and require that animals thermoregulate by moving to terrestrial or fossorial habitats
[@sundayThermalsafetyMarginsNecessity2014] (Fig. \@ref(fig:fig4)). Moreover, low minimum
temperatures reduced positive effects of canopy height on verticality for birds and reptiles (Fig.
\@ref(fig:fig3)). Cold temperatures at higher latitudes can cause ectotherm body temperatures to
exceed critical thermal minima outside fossorial habitats or impose energetic constraints to
endotherms, particularly in the windier conditions that characterize the canopy
[@sundayThermalsafetyMarginsNecessity2014] [REPLACE MAMMAL EXAMPLE WITH BIRD PHYSIOLOGY]. The
physiological constraints imposed by minimum temperatures may act across the entire canopy given
given their shallow vertical gradient, (Fig. \@ref(fig:fig4)), posing severe limitations on the
capacity for taller canopies to increase species richness in boreal forests.

@@

Decades of research have elucidated the mechanisms that allow mountains to generate high levels of biodiversity, particularly in the tropics [CITE]. Yet only recently has the three-dimensional nature of terrestrial landscapes been widely recognized as a critical dimension to our understanding of biogeographic patterns [@fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a]. We bring the vertical dimension of habitats to the forefront of global biogeography by showing that the climatic characteristics of wooded ecosystems from macro- to micro-scales mediate the capacity for tall forests to support high levels of biodiversity via impacts on the vertical structure of vertebrate assemblages. Tropical rainforests, with wet and thermally stable climates that increase the structural complexity of vegetation, reduce physiological stress in the canopy, and produce distinct vertical climate bands, are able to support fine-scale partitioning of resources and climatic niches throughout the entire forest volume. Consequently, taller trees in tropical forests support higher biodiversity and greater increases in species richness than temperate and boreal ecosystems of equivalent heights where climatic conditions reduce the structural complexity of vegetation, pose physiological constraints on arboreal lifestyles, and reduce vertical climate zonation from the ground to the canopy.

@@

```{r fig1, fig.id="fig1"}
# Fig. 1

# Legend: Spatial distribution of a) canopy height, b) vertebrate richness (all vertebrates combined), and
# c) deviance residuals for a negative binomial model of total vertebrate richness by canopy height
# d) Relationships between species richness and canopy height for each taxon in each
# biome category. Dots indicate observed values and lines indicate negative binomial
# model fits of richness by canopy height.

@@

```{r fig2, fig.id="fig2", fig.cap="Global variation in the vertical structure of vertebrate assemblages and its association with species richness for birds, mammals, reptiles, and amphibians. a) The average ± SD proportion of species exhibiting arboreal (red), terrestrial (yellow), and fossorial (blue) strategies across latitudes. Proportions can exceed one because species can exhibit multiple vertical strategies. b) The verticality of vertebrate communities in wooded biomes across the globe. High verticality (red) indicates that the assemblage has a higher vertical niche position than expected by random chance." c) Correlations between verticality and species richness globally. Lines indicate linear best fit line with 95% CIs. }
# Fig. 1

@@

```{r fig4, fig.id="fig4", fig.cap=", fig.width=140, fig.height=120}

@@

To examine spatial patterns of assemblage verticality independently of species richness and accounting for biogeographic history, we calculated the standard effect size (SES) of mean verticality (hereafter ‘verticality’): $SES=(x_i-\bar{x})/sd(x)$ where x is a vector of expected mean verticality scores given random selection of species from a species pool. For each vertebrate class, we defined species pools as all species inhabiting wooded habitats within a given biogeographic realm (Dinerstein et al. 2017) and generated a null distribution of expected verticality for each grid cell based on 100 simulations. Realms included Palearctic, Nearctic, Indomalayan, Neotropic, Oceania, Afrotropic, and Australasia. For each simulation, we randomized verticality within the grid cell by randomly drawing the same number of verticality scores from the respective species pool as the observed richness in the cell. Highly positive SES verticality scores represent assemblages having higher verticality than expected given the species richness, while low SES verticality scores represent assemblages having much lower verticality than expected given the species richness.

@@

Vertebrate richness is expected to increase with increasing canopy height. We first modeled richness as a function of canopy height using a generalized linear model with a Poisson distribution and a log-link. Models were produced for global richness across all vertebrate class, global richness for each vertebrate class, and biome richness for each vertebrate class. Initial model evaluation indicated over-dispersion in model residuals. We therefore fit a negative binomial model with a log-link function. Since richness often exhibits a log-log relationship with covariates, we additionally fit a model of richness as a function of log(canopy height), and compared this model to the original using AIC and pseudo-R^2^ values. We report results of richness as a function of canopy height, as it consistently performed better except for the model for reptiles in temperate forests.

@@

We also examined relationships between richness and verticality using Pearson correlations.

@@

We examined relationships between verticality and environmental covariates using spatial linear mixed models (LMMs). Grid cells with fewer than five species were excluded, because they may provide unrealiable estimates of the SES. To linearize the relationships between predictor and response variables, we log10-transformed precipitation of the driest month. We then scaled all predictor variables to mean zero and variance one and examined their multicollinearity using Pearson correlations. We excluded precipitation seasonality, temperature seasonality, and foliage height diversity due to strong correlations (|r| > 0.8) with dry season precipitation, minimum temperature of the coldest month, and canopy height, respectively. Additionally, we included a quadratic effect for maximum temperature of the warmest month, since extremely high temperatures may inhibit arboreal life strategies.

Spatial LMMs were fit using the R package sdmTMB [@andersonSdmTMBPackageFast2024], which approximates a spatial random effect using a finite element analysis mesh as implemented in R-INLA [@thorsonSpatioTemporalModelsEcologists2024]. The full model can be described as:

$$
\begin{align}
y_s ∼ 𝜇_{s}, \\
𝜇_s = 𝜲_s𝛽 + ⍵_s, \\
⍵_s ∼ MVNormal(0,𝚺_⍵)
\end{align}
$$

where $y_s$ is verticality at location $s$ defined by x and y coordinates, $𝜇_s$ is the expected verticality. $Χ_s$ is a vector of main effect covariates and 𝛽 represents a corresponding vector of estimated main effect coefficients, including additive effects of Tmax~warm~, Tmax~warm~^2^, Tmin~cold~, Temp~diu~, Precip~warm~, log10(Precip~dry~), vegetation density, and canopy height, as well as interactions between canopy height and the five climatic variables. Interactions were included because we expect the impact of canopy height on verticality to decline when harsh macroclimate conditions, including extreme cold, hot, or dry conditions, reduce the suitability of canopy microhabitats for arboreal life strategies, or when high climate variability reduces distinctness of vertical climate bands. $⍵_s$ represents a spatial random effect at location $s$, and describes additional spatial variation in the data not explained by the environmental covariates. The spatial random effect was modeled as a Gaussian random field with a covariance matrix modeled by the Matérn covariance function, which defines the decay rate of correlation with distance. Gaussian random fields were approximated for models of verticality for each taxa using a spatial mesh optimized such that the edge length was as large as possible to reduce computational time while remaining less than five times the spatial range of the model [@andersonSdmTMBPackageFast2024; @krainskiAdvancedSpatialModeling2019].

We validated models first using the ‘sanity’ function from the sdmTMB package. We additionally calculated quantile residuals using the DHARMa R package [@hartigDHARMaResidualDiagnostics2022] to assess residual normality and lack of patterning across space.

@@

To examine vertical microclimate gradients, we mechanistically modeled microclimates from the ground to the canopy using the ‘micropoint’ R package [@macleanMicropointCanopyPoint2025]. Models were produced for boreal forests, temperate forests, and tropical forests from the Americas and from Eurasia.

@@

Modeling microclimates within the canopy requires detailed information on the vertical distribution of vegetation, which we obtained from spaceborne lidar data acquired through NASA's GEDI. The footprint level L2b data was preprocessed by Burns et al. [-@burnsGriddedGEDIVegetation2024; @burnsMultiresolutionGriddedMaps2024] to filter out low quality points and gridded into 1° x 1° tiles. We selected six tiles in which we modeled microclimates, representing boreal, temperate, and tropical forests in the Americas and Eurasia in non-mountainous regions that contained relatively large areas of tree cover based on Google satellite imagery and that were within the latitudinal bounds of the GEDI data. As the highest latitude is 51.6°N, we were only able to produce microclimate models for southern boreal forests. Boreal forests included regions in Canada (50.75°,-89.54°) and Russia (50.8°,137.77°); temperate forests included regions in the midwestern United States (37.69°,-91.08°) and Europe (48.33°,10.50°); tropical forests included regions in northern Peru (-5.14°,-75.57°) and Malaysia (4.95°,117.69°).

Points that fell in areas outside closed forest, as identified by Copernicus Global Dynamic Land Cover 2019 v3 [@buchhornCopernicusGlobalLand2020], areas that experienced deforestation between 2001 and 2024, as identified by Hansen global forest watch [@hansenHighResolutionGlobalMaps2013], and areas above 500 m in elevation were excluded from analysis. We restricted analysis to low elevation forests to control for any elevational impacts on vertical climate gradients. From the remaining points, we identified those that were recorded during the months with the average coldest minimum temperature and average warmest maximum temperature for the year in which they were obtained. These months were identified using ERA5 climate data [@hersbachERA5HourlyData2023]. For minimum and maximum temperature subsets, we selected up to 100 GEDI points within each 1° x 1° tile. However, there were not 100 points that matched our selection criteria in temperate forests during minimum temperature months because GEDI only produces reliable vertical profiles during leaf-on conditions (cite) (26 points in temperate USA and 14 points in temperate Europe). All temperate forest points therefore fell in mixed rather than deciduous forests. There were also only 65 points that met our selection criteria in temperate European forests during the warmest month.

For each point, GEDI data provided information on canopy height, plant area index (PAI), and vertical distribution of the PAI. The GEDI data describes the vertical distribution of PAI as the cumulative PAI from the top of the canopy to the stratum representing the vertical space from 0-5 m above the ground. For example, the PAI 0-5 m includes nearly all of the canopy space and is therefore nearly equal to the total PAI in most cases, while PAI 10-15 m describes the PAI from the top of the canopy to 10-15 m above the ground. However, the microclimate models require that the vertical PAI distribution represent the PAI within individual height bins and that the sum of the vertical PAI profile equal the total PAI. We therefore subtracted total PAI – PAI 0-5 m to represent the PAI in the 0-5 m bin, PAI 0-5 m – PAI 5-10 m to represent the PAI in the 5-10 m bin, and so forth.

We used ERA5 for macroclimate forcing data. To reduce gridding effects of the data, we averaged the four closest grid cells to the location of the GEDI point using inverse distance weighting. ERA5 data was processed into the format required by the model using the 'mcera5' R package [@klingesMcera5DrivingMicroclimate].
 [DETAILS ON DIURNAL TEMPERATURE CORRECTION AND LAND SEA MASK]

### Soil data

Soil data was obtained from the soilgrids database at a 250 m resolution using the soilDB R package, and included sand, silt, and clay content at the top soil layer (0-5 cm below the ground). [DESCRIBE METHODS USED TO CONVERT TO SOIL DATA USED BY MICROCLIMATE MODEL].

@@

```{r richness-mods-results, tab.id="richness-mod-results", tab.lp="stab"}
# table from: "scripts/03_analyses/01_canopyheight_richness_relationships.R"

fit.global.tidy %>% 
  mutate(term = ifelse(term == "canopy_height2", "canopy height", term)) %>% 
  mutate(across(.cols = estimate:p.value, .fns = function(x) {round(x,2)})) %>% 
  rename(SE = std.error, "p-value" = p.value) %>% 
  flextable()

@@

```{r vert-model-results, tab.id="vert-model-results", tab.cap="Coefficient estimates for standardized predictor variables and pseudo-R^2^ values for each model."}

@@

@@

# SCRATCH SPACE

"The observer who leaves the center of the earth by an infinitely small amount compared to the
radius can reach a new world, so to speak, and he can observe more variations in the aspect of the
soil and more modifications in the atmosphere than he would if he were to pass from one latitude to
another" - Humbolt pg 77 Essay on the Geography of Plants

"Current hypotheses for large-scale geographical patterns of species diversity fail to explain the
extraordinarily high biodiversity of mountain regions, particularly in the tropics (2–5). We call
this challenge “Humboldt’s enigma.”" - Rahbek

-   TRY RELATIVIZING RATES OF INCREASE BECAUSE THERE ARE MORE SPECIES IN THE TROPICS. EG. INCREASE
    IN 1 SPECIES OUT OF 10 TOTAL SPECIES = INCREASE IN 10 SPECIES IF THERE ARE 100 SPECIES IN THE
    REGION
-   on second thought, I don't think this is necessary because we want to know why more species
    can't exist in trees in boreal regions

Just as topographic complexity and climate explain Humbolt's enigma in mountains, vegetative
complexity and climate may play key roles in mediating the relationship between canopy height and
species richness by impacting the availability of vertical niche opportunities and physiologically
constraining arboreal lifestyles [CITE]. Densely layered forest structures not only provide habitat
across the forest profile, but produce vertical microclimate gradients via interactions with
macroclimate conditions, solar radiation, and wind speed. From the humid and thermally stable
understory to the dry and thermally variable canopy that experiences both hot and cold temperature
extremes [CITE], this vertical gradient may interact with latitudinal climate gradients to help or
hinder the capacity for tall forests to generate high levels of biodiversity. For example, increased
exposure to extreme cold in the canopies of high latitude forests may truncate the volume of habitat
that can support arboreal species and thus limit richness despite tall canopies
[@oliveiraVerticalStratificationInfluences2019]. Moreover, relatively low thermal variability in the
canopies of tropical forests may increase the distinctness of vertical climate bands that support
high turnover and thus enhance the capacity for tall forests to generate biodiversity in the tropics
[CITE].

in wet tropical forests, macroclimate conditions produce warmer and wetter canopy microclimates,
which reduce thermal and hydric stress and may allow species to finely partition the entire
three-dimensional habitat space [@oliveiraVerticalStratificationInfluences2019]. Taller canopies in
tropical regions could thus support higher animal verticality and richness. In contrast, extreme
cold, or the combination of heat and drought may impede the evolution of arboreal lifestyles,
constraining biodiversity to lower vertical positions tat are more climatically buffered (cite).
Harsh macroclimate conditions may therefore truncate the availability of vertical habitat space,
thereby limiting vertical stratification and species richness despite the presence of vegetation.

Integrating new technologies can shed light on the mechanisms driving decades-old ecological
patterns.

But only within the last decade have we been able to measure forest structure and acquire data on
vertebrate use of canopy strata at a global extent, allowing us to now gain a comprehensive
understanding of the impact of vertical habitat structure on species richness patterns.

In forests, species assemblages stratify vertically along resource and microclimate gradients that
extend from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a;
@xingEcologicalPatternsProcesses2023a]. Tall, multi-layered forests increase both habitat volume and
niche opportunities, enabling species to partition habitat across vertical strata akin to
multi-floored apartments [@macarthurLimitingSimilarityConvergence1967;
@macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023].

>>>>

---
bibliography: "r-refs.bib"
csl: science.csl
output: 
  officedown::rdocx_document:
    reference_docx: "style-guide.docx"
    mapstyles:
      Normal: ['First Paragraph']
plots:
  caption:
    style: Image Caption
    pre: 'Figure '
    sep: ': '
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \rhead{Global Verticality}
- \begin{document}
---

@@

Maximum and minimum temperatures had less consistent effects across taxa (Fig. \@ref(fig:fig3)), but may still place physiological constraints on arboreal life strategies that influence the capacity for taller canopies to support richer species assemblages, particularly in regions exhibiting steeper vertical gradients characterized by hotter maximum and colder minimum temperatures. For example, high maximum temperatures generally reduced verticality and/or reduced the positive impact of canopy height on verticality, barring mammals for which verticality responded positively to maximum temperatures. High temperatures may cause ectotherm body temperatures to exceed critical thermal limits [@sundayThermalsafetyMarginsNecessity2014]. While consistent water availability in the canopies of tropical wet forests may allow ectotherms to maintain body temperature below critical limits, the drier conditions of temperate forests combined with steeper vertical gradients in maximum temperatures and canopy temperatures that exceed 45°C may impede the use of arboreal habitats and require that animals thermoregulate by moving to terrestrial or fossorial habitats [@sundayThermalsafetyMarginsNecessity2014] (Fig. \@ref(fig:fig4)). Moreover, while models indicated that higher minimum temperatures were associated with lower verticality for birds and amphibians because verticality is low in hot but drier tropical regions while remaining high in cooler but wet tropical montane regions, low minimum temperatures reduced positive effects of canopy height on verticality for birds and reptiles (Fig. \@ref(fig:fig3)). The interaction with canopy height reflects the physiological constraints that cold temperatures can impose on arboreal life strategies. Cold temperatures at higher latitudes can cause ectotherm body temperatures to exceed critical thermal minima outside fossorial habitats or impose energetic constraints on endotherms, particularly in the windier conditions of the upper canopy [@sundayThermalsafetyMarginsNecessity2014; @villen-perezWoodlandMediterraneanBirds2014; @wolfThermalEffectsRadiation1996]. For example, birds that typically forage in the upper canopy shift to lower canopy strata during periods of cold winter temperatures [@villen-perezWoodlandMediterraneanBirds2014]. Cold temperatures therefore limit the degree to which animals can specialization to upper canopy strata and consequently prevent taller canopies from increasing species richness.

@@

Centuries of research have elucidated the mechanisms that allow mountains to generate high levels of biodiversity, particularly in the tropics [@vonhumboltEssayGeographyPlants2013; @janzenWhyMountainPasses1967; @rahbekHumboldtsEnigmaWhat2019]. Yet only recently has the three-dimensional nature of terrestrial landscapes been widely recognized as a critical dimension to our understanding of biogeographic patterns [@fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a]. We bring the vertical dimension of habitats to the forefront of global biogeography by showing that the climatic characteristics of wooded ecosystems from macro- to micro-scales mediate the capacity for tall forests to support high levels of biodiversity via impacts on the vertical structure of vertebrate assemblages. Tropical rainforests, with wet and thermally stable climates that increase the structural complexity of vegetation, reduce physiological stress in the canopy, and produce distinct vertical climate bands, are able to support fine-scale partitioning of resources and climatic niches throughout the entire forest volume. Consequently, taller trees in tropical forests support higher biodiversity and greater increases in species richness than temperate and boreal ecosystems of equivalent heights where climatic conditions reduce the structural complexity of vegetation, pose physiological constraints on arboreal lifestyles, and reduce vertical climate zonation from the ground to the canopy.

@@

```{r fig1, fig.id="fig1", fig.cap="Spatial distribution of a) canopy height, b) vertebrate richness (for all vertebrates), and c) deviance residuals for a negative binomial model of total vertebrate richness by canopy height. d) Relationships between species richness and canopy height for each taxon in different biomes. Dots indicate observed values and lines indicate negative binomial model fits of richness as a function of the interaction between canopy height and biome.", fig.height=100, fig.width=180}

# Fig. 1

@@

```{r fig2, fig.id="fig2", fig.cap="Global variation in the vertical structure of vertebrate assemblages and its association with species richness for birds, mammals, reptiles, and amphibians. a) The average ± SD proportion of species exhibiting arboreal (red), terrestrial (yellow), and fossorial (blue) strategies across latitudes. Proportions can exceed one because species can exhibit multiple vertical strategies. b) The verticality of vertebrate communities in wooded biomes across the globe. High verticality (red) indicates that the assemblage has a higher vertical niche position than expected by random chance." c) Correlations between verticality and species richness globally. Lines indicate linear best fit line with 95% CIs., fit.width = 150, fig.height=150}

# Fig. 2

@@

```{r fig4, fig.id="fig4", fig.cap="a) Locations of modeled vertical microclimate profiles. b) Average hourly microclimate across relative heights in the canopy. Relative height is calculated as absolute height above ground divided height of the canopy. Shading represents the range of temperatures experienced in the uppermost vertical band (0.9-1 relative height) and the lowermost vertical band (0-0.1 relative height). c) Vertical profiles of microclimate variables and plant area volume density (PAVD) (median +/- SD) in tropical, temperate, and boreal forest sites. PAVD was obtained from GEDI shots.", fig.width=140, fig.height=120}

@@

To examine spatial patterns of assemblage verticality independently of species richness and accounting for biogeographic history, we calculated the standard effect size (SES) of mean verticality (hereafter ‘verticality’): $SES=(x_i-\bar{x})/sd(x)$ where x is a vector of expected mean verticality scores given random selection of species from a species pool. For each vertebrate class, we defined species pools as all species inhabiting wooded habitats within a given biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality for each grid cell based on 100 simulations. Realms included Palearctic, Nearctic, Indomalayan, Neotropic, Oceania, Afrotropic, and Australasia. For each simulation, we randomized verticality within the grid cell by randomly drawing the same number of verticality scores from the respective species pool as the observed richness in the cell. Highly positive SES verticality scores represent assemblages having higher verticality than expected given the species richness, while low SES verticality scores represent assemblages having much lower verticality than expected given the species richness.

@@

Vertebrate richness is expected to increase with increasing canopy height. We first modeled richness as a function of canopy height using a generalized linear model with a Poisson distribution and a log-link. Models were produced for global richness across all vertebrate classes and global richness for each vertebrate class. For each vertebrate class, we then modeled richness as a function of the interactive effect between canopy height and biome using a means parameterization. Initial model evaluation indicated over-dispersion in model residuals. We therefore fit negative binomial models with a log-link function. Since richness often exhibits a log-log relationship with covariates, we additionally fit models of richness as a function of log(canopy height) or the interactive effect of log(canopy height) and biome, and compared this model to the original using AIC and conditional percent deviance explained (CPDE). Models fit with canopy height rather than log(canopy height) consistently performed better, and we report these results.

@@

We additionally examined relationships between richness and verticality using Pearson correlations.

@@

We examined relationships between the verticality of each taxon and environmental covariates described above using spatial linear mixed models (LMMs). Grid cells with fewer than five species were excluded, because they may provide unreliable estimates of the SES. To linearize the relationships between predictor and response variables, we log10-transformed precipitation of the driest month. We then scaled all predictor variables to mean zero and variance one and examined their multicollinearity using Pearson correlations and variance inflation factors. We excluded precipitation seasonality, temperature seasonality, and foliage height diversity due to strong correlations (|r| > 0.8) with dry season precipitation, minimum temperature of the coldest month, and canopy height, respectively. The remaining variables all had VIF < 5. Additionally, we included a quadratic effect for maximum temperature of the warmest month, because extremely high temperatures are expected to inhibit arboreal life strategies as well as interaction effects between the climatic covariates and canopy height. Interactions were included because we expect the impact of canopy height on verticality to decline when harsh macroclimate conditions, including extreme cold, hot, or dry conditions, reduce the suitability of canopy microhabitats for arboreal life strategies, or when high climate variability reduces distinctness of vertical climate bands. The fixed effects therefore included Tmax~warm~, Tmax~warm~^2^, Tmin~cold~, Temp~diu~, Precip~warm~, log10(Precip~dry~), vegetation density, and canopy height, as well as interactions between canopy height and the climatic covariates.

Spatial LMMs were fit using the R package sdmTMB [@andersonSdmTMBPackageFast2024], which accounts for spatially autocorrelated data using stochastic partial differential equations (SPDE) to approximate spatial Gaussian random fields as Gaussian Markov Random Fields (GMRF) with a Matérn covariance function [@lindgrenExplicitLinkGaussian2011]. The full model can be described as:

$$
\begin{align}
y_s ∼ N(𝜇_{s}, \sigma^2), \\
𝜇_s = 𝜲_s𝛽 + ⍵_s, \\
⍵_s ∼ MVNormal(0,𝚺_⍵),
\end{align}
$$

where the response variable at location $s$ is normally distributed with mean $\mu_s$ and variance $\sigma^2$; $𝜇_s$ is the expected verticality. $Χ_s$ is a design matrix at location $s$; 𝛽 is a vector of fixed-effect coefficients; $⍵_s$ represents a spatial random effect at location $s$; and $\Sigma_{\omega}$ represents the covariance matrix of the spatial random field. The SPDE method requires a finite element analysis mesh, which consists of a set of triangles covering the spatial domain, and allows for the approximation of the spatial random effect at mesh vertices [@thorsonSpatioTemporalModelsEcologists2024]. We created a mesh for each vertebrate class using *fmesher* and tuned each mesh, such that the maximum edge length was as large as possible to reduce computational time while remaining less than five times the spatial range of the model [@andersonSdmTMBPackageFast2024; @krainskiAdvancedSpatialModeling2019].

We evaluated models using the sanity check function in sdmTMB and by examining quantile residuals using the DHARMa R package [@hartigDHARMaResidualDiagnostics2022]. To determine the degree to which environmental factors improve model performance, we compare this model to a spatial random effect model with  model with We also report conditional percent deviance explained (CPDE).

@@

To examine vertical microclimate gradients, we mechanistically modeled microclimates from the ground to the canopy using the ‘micropoint’ R package [@macleanMicropointCanopyPoint2025; @macleanMicroclimfFastModelling2025]. Point models of vertical climate profiles were produced for boreal, temperate, and tropical forests from the Americas and Eurasia within six 1° x 1° tiles located in regions with relatively flat topography that contained relatively large areas of tree cover based on Google satellite imagery and that were within the latitudinal bounds of the GEDI data, which provided information on vegetation structure (see below). Boreal forests included regions in Canada (50°,-90°) and Russia (50°,137°); temperate forests included regions in the midwestern United States (37°,-92°) and Europe (48°,10°); tropical forests included regions in northern Peru (-6°,-76°) and Malaysia (4°,117°), where coordinates indicate the lower left corners of tiles.

Points within these tiles were selected as locations of GEDI footprint level L2b data, which was preprocessed by Burns et al. [-@burnsGriddedGEDIVegetation2024; @burnsMultiresolutionGriddedMaps2024] to filter out low quality points. From the high quality points, we excluded points that 1) fell outside closed forest, as identified by Copernicus Global Dynamic Land Cover 2019 v3 [@buchhornCopernicusGlobalLand2020], 2) were located in areas that experienced deforestation between 2001 and 2024, as identified by Hansen global forest watch [@hansenHighResolutionGlobalMaps2013], 3) were obtained from coverage rather than power beams, 4) represented vegetation less than 10 m tall, and 4) occurred above 500 m in elevation. We restricted analysis to low elevation forests to control for any elevational impacts on vertical climate gradients. From the remaining points, we identified those that were recorded during the months with the average coldest minimum temperature and average warmest maximum temperature for the year in which they were obtained. These months were identified using the ECMWF ERA5 reanalysis product [@hersbachERA5HourlyData2023]. For minimum and maximum temperature subsets, we selected up to 100 GEDI points within each 1° x 1° tile. However, there were only 26 points in temperate USA and 14 points in temperate Europe during minimum temperature months because GEDI only produces reliable vertical profiles during leaf-on conditions (cite). The points retained therefore represented mixed rather than deciduous forest types. Furthermore, only 65 points met our selection criteria in temperate European forests during the warmest month. For each point, the microclimate model was run for the month and year in which the point was obtained at 0.15 cm above ground, 2 m above ground, and then from 5 m to the top of the canopy at 5 m intervals (equivalent to the vertical resolution of GEDI data) to produce hourly estimates of air temperature and relative humidity.

The ECMWF ERA5 reanalysis product [@hersbachERA5HourlyData2023] was used for meteorological inputs to the model, which provided near surface, open-air climate conditions at a 0.25$^\circ$ spatial resolution and hourly temporal resolution. ERA5 climate data was processed using the *mcera5* R package, which converts ERA5 variables into those required by the model and applies a coastal diurnal temperature correction to terrestrial locations located within predominanrly marine grid cells [@klingesMcera5DrivingMicroclimate2022]. Meteorological conditions at the GEDI point of interest were then extracted from the climate grid by taking the weighted average of climate conditions across the four nearest neighboring cells, where the weight was defined as the inverse distance to the point. Finally, we applied an elevational correction that adjusts temperature and pressure at the location of interest by accounting for the difference in elevation between the center of the ERA5 grid cell and the point of interest using a humidity-dependent lapse rate.

@@

Vegetation structure influences the amount of radiation and vertical air mixing. We obtained information on vegetation from the GEDI footprint level L2b data.
For each point, GEDI data provided information on canopy height (RH100; the height at which 100% of the waveform was returned), plant area index (PAI), and the vertical distribution of the PAI. The GEDI data describes the vertical distribution of PAI as the cumulative PAI from the top of the canopy to the stratum representing the vertical space from 0-5 m above the ground. For example, the PAI 0-5 m includes nearly all of the canopy space and is therefore nearly equal to the total PAI in most cases, while PAI 10-15 m describes the PAI from the top of the canopy to 10-15 m above the ground. However, the microclimate models require that the vertical PAI distribution represent the PAI within individual height bins and that the sum of the vertical PAI profile equal the total PAI. We therefore subtracted total PAI – PAI 0-5 m to represent the PAI in the 0-5 m bin, PAI 0-5 m – PAI 5-10 m to represent the PAI in the 5-10 m bin, and so forth.

*Micropoint* requires a number of additional parameters describing vegetation structure that are not available from remote sensing products, such leaf angle distribution and stomatal conductance, among others. We use *microclimf's* 'vegpfromhab()' function to estimate these parameters based on land cover type. Land cover was determined based on MODIS land cover type (MCD12Q1) version 6.1 for the year 2022.

Vegetation parameters were assumed to remain constant for the month and year in which the GEDI shot was obtained.

[WE SHOULD PRODUCE MODELS AT DRIEST MONTH OF THE YEAR]

### Topography and soil composition

To characterize topography, we extracted elevation from the lowest mode return of the GEDI shot. Slope and aspect were calculated using *terrain()* based on SRTM elevation data at a 1 km resolution, which was obtained using the *geodata* R package. Sand, silt, and clay content in the top soil layer (0-5 cm below the ground) were obtained from the SoilGrids 2.0 at a 250 m resolution using the soilDB R package, and included sand, silt, and clay content at the top soil layer (0-5 cm below the ground). Soil type was determined based on the relative percentage of sand, silt, and clay based on USDA soil classifications using the *soiltexture* R package. Soil properties were then defined for the given soil class as default values provided by *micropoint*.

### Microclimate analysis

To examine global patterns of vertical microclimate profiles, we calculated the average minimum temperature of the coldest month, maximum temperature of the warmest month, minimum relative humidity of the warmest month, and diurnal temperature range of the coldest month. For each forest type (boreal, temperate, and tropical), we graphed the median climate value within vertical intervals, each representing 10% of the total canopy height. Therefore, values at the top of the canopy always occur in the 0.9-1 relative height interval regardless of the absolute height of the canopy. We additionally examined diurnal temperature fluctuations across forest types by taking the mean value across points for a given relative height and hour of the day.

### Software
All analyses were conducted using R Statistical Software (v4.4.1; @rcite).

@@

```{r richness-mods-results, tab.id="richness-mod-results", tab.lp="stab". tab.cap="Coefficient estimates (with 95% confidence intervals) and conditional percent deviance explained (CPDE) for global models of richness as a function of canopy height for all taxa and each vertebrate class and for biome models of richness as a fucntion of canopy height x biome. Biome models used a means parameterization such that the effect of biome represents the mean estimated richness at a canopy height of zero and the interaction effect represents the estimated impact of canopy height on richness for each biome."}

# tables from: "scripts/03_analyses/01_canopyheight_richness_relationships.R"

tidy.global = fit.global.tidy %>% 
  mutate(term = ifelse(term == "canopy_height2", "Canopy height", term),
         taxa = "All vertebrates") %>% 
  mutate(across(.cols = estimate:conf.high, .fns = function(x) {round(x,2)})) %>% 
  mutate(estimate = ifelse(term != "CPDE (%)",
                           paste0(estimate, " (", conf.low, ", ", conf.high, ")"),
                           paste0(estimate))) %>% 
  dplyr::select(taxa, term, estimate) %>% 
  pivot_wider(names_from = "taxa", values_from = "estimate")

tidy.global.taxa = global.taxa.tidy %>% 
  dplyr::select(taxa, term, estimate, conf.low, conf.high) %>% 
  mutate(term = ifelse(term == "canopy_height2", "Canopy height", term)) %>%
  mutate(across(.cols = estimate:conf.high, .fns = function(x){round(x,3)})) %>% 
  arrange(taxa) %>% 
  mutate(estimate = ifelse(term != "CPDE (%)",
                           paste0(estimate, " (", conf.low, ", ", conf.high, ")"),
                           paste0(estimate)),
         taxa = factor(taxa, levels = c("Birds", "Mammals", "Reptiles", "Amphibians"))) %>% 
  dplyr::select(taxa, term, estimate) %>% 
  pivot_wider(names_from = "taxa", values_from = "estimate")

tidy.global = left_join(tidy.global, tidy.global.taxa, by = "term") %>% 
  mutate(model = "Global model") %>% 
  relocate(model)

tidy.biome = biome.mod.tidy %>% 
  dplyr::select(taxa, term, estimate, conf.low, conf.high) %>% 
  mutate(term = gsub("e2", "e = ", term),
         term = gsub(":", " x ", term),
         term = gsub("2", "", term)) %>% 
  mutate(across(.cols = estimate:conf.high, .fns = function(x){round(x,3)})) %>% 
  mutate(estimate = ifelse(term != "CPDE (%)",
                           paste0(estimate, " (", conf.low, ", ", conf.high, ")"),
                           paste0(estimate)),
         taxa = factor(taxa, levels = c("Birds", "Mammals", "Reptiles", "Amphibians"))) %>% 
  dplyr::select(taxa, term, estimate) %>% 
  pivot_wider(names_from = "taxa", values_from = "estimate") %>% 
  mutate(model = "Biome model") %>% 
  relocate(model)

tidymods = bind_rows(tidy.global, tidy.biome)

tidymods %>% 
  flextable() %>% 
  hline(i = 3, part = "body", border = fp_border(color = "black", width = 3)) %>% 
  merge_v(j = c("model")) %>% 
  autofit()

@@

```{r vert-model-results, tab.id="vert-model-results", tab.lp="stab", tab.cap="Coefficient estimates for standardized predictor variables and pseudo-R^2^ values for each model."}

@@

```{r splot-microtempdiu, fig.id="microtempdiu", fig.lp="suppfig", width = 120, height = 70, fig.cap="Average hourly microclimate across relative heights in the canopy. Relative height is calculated as absolute height above ground divided height of the canopy."}

include_graphics("figures/ms_figures/supp_figs/microclim_temp_diu_mintempMonth.png", error = F)

@@
